---
title: "ADA Project"
author: "Alex Knitter"
date: '2022-09-28'
output: html_document
---

# Project plan

#### 1. Research question

Among non-elderly adults with asthma in 2021, how did daily fruit and vegetable intake affect their reported number of days of poor physical and mental health?

#### 2. Identify each PICOT item in your research question

*Population*: individuals 18-64 years of age with asthma

*Intervention*: daily a) fruit and b) vegetable intake

*Comparison group*: those with a) fruit and b) vegetable intake less than once per day

*Outcome*: Self-reported number of days when poor physical or mental health limited usual activities (in the past 30 days)

*Time frame*: 2021

#### 3. Data source

CDC, BRFSS 2021 (<https://www.cdc.gov/brfss/annual_data/annual_2021.html>)

#### 4. Variables

-   Age
-   Sex
-   Race/ethnicity
-   Income
-   Residence in urban/rural area
-   Daily fruit intake (yes/no)
-   Daily vegetable intake (yes/no)
-   Days in past month when poor physical or mental health limited usual activities



### Importing the Data

Dataset loaded directly from CDC website: <https://www.cdc.gov/brfss/annual_data/2021/files/LLCP2021XPT.zip>

```{r load-packages}
pacman::p_load(haven, tidyverse, table1, labelled, magrittr, DiagrammeR, lmtest, MASS)
```

```{r import-data}
temp <- tempfile(fileext = ".zip")
download.file(url = "https://www.cdc.gov/brfss/annual_data/2021/files/LLCP2021XPT.zip", destfile = temp)
brfss2021 <- read_xpt(file = temp)

# if importing data from web link is giving difficulty, run the below to import from files
# brfss2021 <- read_xpt("LLCP2021.XPT ")
```


### Preparing the Data

```{r clean-data}
brfss2021Clean <- brfss2021 %>%
  # select variables of interest
  dplyr::select(ASTHNOW, POORHLTH, `_FRTLT1A`, `_VEGLT1A`, `_INCOMG1`, `_URBSTAT`, `_RACE`, `_SEX`, `_AGEG5YR`, PHYSHLTH, MENTHLTH, `_FRT16A`, `_VEG23A`) %>%
  # rename variables for ease of use
  rename(has_asthma = ASTHNOW) %>%
  rename(days_poor_health = POORHLTH) %>%
  rename(fruit_daily = `_FRTLT1A`) %>%
  rename(veg_daily = `_VEGLT1A`) %>%
  rename(income = `_INCOMG1`) %>%
  rename(urban_rural = `_URBSTAT`) %>%
  rename(race_eth = `_RACE`) %>%
  rename(sex = `_SEX`) %>%
  rename(age_cat = `_AGEG5YR`) %>%
  rename(days_physical_bad = PHYSHLTH) %>%
  rename(days_mental_bad = MENTHLTH) %>%
  rename(fruit_high = `_FRT16A`) %>%
  rename(veg_high = `_VEG23A`) %>%
  # recode numeric variable values
  mutate(days_poor_health = as.numeric(recode(as.character(days_poor_health), 
                                        '88' = '0', # none, 0 days poor health
                                        '77' = NA_character_, # don't know
                                        '99' = NA_character_))) %>% # refused
  mutate(days_physical_bad = as.numeric(recode(as.character(days_physical_bad), 
                                        '88' = '0', # none, 0 days poor health
                                        '77' = NA_character_, # don't know
                                        '99' = NA_character_))) %>% # refused
  mutate(days_mental_bad = as.numeric(recode(as.character(days_mental_bad), 
                                        '88' = '0', # none, 0 days poor health
                                        '77' = NA_character_, # don't know
                                        '99' = NA_character_))) %>% # refused
  # recode factor variable levels
  # for most variables, 9 = don't know, refused, missing
  ## for has_asthma, 7 = don't know, 9 = refused
  ## for age_cat, 14 = don't know, refused, missing
  mutate(has_asthma = recode_factor(has_asthma,
                                    '1' = "Has asthma",
                                    '2' = "No current asthma",
                                    '7' = NA_character_,
                                    '9' = NA_character_)) %>%
  mutate(fruit_daily = recode_factor(fruit_daily,
                                     '1' = "Consumed fruit at least once daily",
                                     '2' = "Consumed fruit < once daily",
                                     '9' = NA_character_)) %>%
  mutate(veg_daily = recode_factor(veg_daily,
                                     '1' = "Consumed vegetables at least once daily",
                                     '2' = "Consumed vegetables < once daily",
                                     '9' = NA_character_)) %>%
  mutate(income = recode_factor(income,
                                '1' = "Less than $15,000",
                                '2' = "$15,000 to < $25,000",
                                '3' = "$25,000 to < $35,000",
                                '4' = "$35,000 to < $50,000",
                                '5' = "$50,000 to < $100,000",
                                '6' = "$100,000 to < $200,000",
                                '7' = "$200,000 or more",
                                '9' = NA_character_)) %>%
  mutate(urban_rural = recode_factor(urban_rural,
                                     '1' = "Urban",
                                     '2' = "Rural")) %>%
  mutate(race_eth = recode_factor(race_eth,
                                '1' = "White non-Hispanic",
                                '2' = "Black non-Hispanic",
                                '3' = "American Indian / Alaskan Native non-Hispanic",
                                '4' = "Asian non-Hispanic",
                                '5' = "Native Hawaiian / Pacific Islander non-Hispanic",
                                '6' = "Other race non-Hispanic",
                                '7' = "Multiracial non-Hispanic",
                                '8' = "Hispanic",
                                '9' = NA_character_)) %>%
  mutate(age_cat = recode_factor(age_cat,
                                '1' = "18-24 years",
                                '2' = "25-29 years",
                                '3' = "30-34 years",
                                '4' = "35-39 years",
                                '5' = "40-44 years",
                                '6' = "45-49 years",
                                '7' = "50-54 years",
                                '8' = "55-59 years",
                                '9' = "60-64 years",
                                '10' = "65+ years",
                                '11' = "65+ years",
                                '12' = "65+ years",
                                '13' = "65+ years",
                                '14' = NA_character_)) %>%
  mutate(sex = recode_factor(sex,
                                     '1' = "Male",
                                     '2' = "Female"))

summary(brfss2021Clean)
# frequencies are consistent with BRFSS codebook
```


```{r data-trim}

# include only participants meeting inclusion criteria
# count the number of participants lost with each criteria applied
# assign() stores the dataset after each filter step

brfssCleanFull <- brfss2021Clean %>%
  # include only those with current asthma
  filter(has_asthma == "Has asthma") %>%
  # n = 42660 in temp dataframe after filtering by asthma
  assign("sub_asthma", ., envir = .GlobalEnv) %>%
  # exclude elderly adults, over 65 years of age
  filter(age_cat != "65+ years") %>%
  # n = 28912 after filtering by asthma and age
  assign("sub_asthma_age", ., envir = .GlobalEnv) %>%
  # include only those with valid daily fruit and veg intake
  #   defined by CDC, Fruit >16 per day or Vegetables >23 per day
  filter(fruit_high==1, veg_high==1) %>%
  # n = 28618 after filtering by asthma age and daily f/v intake
  assign("sub_asthma_age_fv", ., envir = .GlobalEnv) %>%
  # drop these filtering variables from dataset
  #   not needed further for analyses
  dplyr::select(-has_asthma, -fruit_high, -veg_high, -days_physical_bad, -days_mental_bad) %>%
  droplevels()

nrow(brfss2021Clean)-nrow(sub_asthma)
nrow(sub_asthma)-nrow(sub_asthma_age)
nrow(sub_asthma_age)-nrow(sub_asthma_age_fv)
nrow(sub_asthma_age_fv)-nrow(brfssCleanFull)

# take a look!
summary(brfssCleanFull)
```


### Explore the dataset

```{r check-missing}
# check if NAs are highly missing for certain patterns/pairs
mice::md.pattern(brfssCleanFull, rotate.names=TRUE)

mice::md.pairs(brfssCleanFull)

# some variables with high Ns missing
# but no clear patterns, seem proportional

# To note: participants were not asked about poor health days if they reported no poor a) physical or b) mental health days in prior questions
## if daily fruit/veg were protective, excluding these participants would bias findings towards the null
```



```{r final-data}
# handling missingness by performing complete case analysis
# therefore, exclude observations with missing values
# count the number of participants lost with each criteria applied
# assign() stores the dataset after each filter step
brfss <- brfssCleanFull %>%
  drop_na(days_poor_health) %>%
  assign("sub_na1_healthdays", ., envir = .GlobalEnv) %>%
  drop_na(fruit_daily) %>%
  assign("sub_na2_fruit", ., envir = .GlobalEnv) %>%
  drop_na(veg_daily) %>%
  assign("sub_na3_veg", ., envir = .GlobalEnv) %>%
  drop_na(income) %>%
  assign("sub_na4_income", ., envir = .GlobalEnv) %>%
  drop_na(urban_rural) %>%
  assign("sub_na5_rural", ., envir = .GlobalEnv) %>%
  drop_na(race_eth) %>%
  assign("sub_na6_race", ., envir = .GlobalEnv) %>%
  drop_na(sex) %>%
  assign("sub_na7_sex", ., envir = .GlobalEnv) %>%
  drop_na(age_cat) %>%
  assign("sub_na8_age", ., envir = .GlobalEnv)

# Note: this should not be used in place of missing n for each variable, exclusion steps are cumulative
## ie, individuals with missing race may have been excluded in an earlier step
```




```{r label-variables}
# add labels to all variables

label(brfss$days_poor_health) <- "Days of Poor Health in the Past 30 Days"
label(brfss$fruit_daily) <- "Daily Fruit Consumption in the Past 30 Days"
label(brfss$veg_daily) <- "Daily Vegetable Consumption in the Past 30 Days"
label(brfss$age_cat) <- "Age"
label(brfss$sex) <- "Sex"
label(brfss$race_eth) <- "Race/Ethnicity"
label(brfss$urban_rural) <- "Urban/Rural Status"
label(brfss$income) <- "Income Category"
```


### Flow diagram

```{r figure1}
#Figure 1
grViz("digraph flowchart {

      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle, fontsize=15] 
      #notes indicate how many boxes you will have in your diagram. Since I have two sets of exclusions, I will have three nodes.
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab7 [label = '@@7']
      tab8 [label = <
      <b>@@8</b>
      >]
      
      # edge definitions with the node IDs are used to indicate how the rectangle boxes flow from each other. This is a simple flow chart but I could create branching instead as below
      tab1 -> tab2 
      tab1 -> tab3
      tab3 -> tab4
      tab3 -> tab5
      tab5 -> tab6
      tab5 -> tab7
      tab5 -> tab8
}
      #This set of code provides the text in each rectangle box.
      [1]: 'BRFSS 2021 survey participants N= 438,693'
      [2]: 'Excluding 396,033 individuals \\n without current asthma'
      [3]: 'Individuals with current asthma n= 42,660'
      [4]: 'Excluding 13,748 individuals over age 65'
      [5]: 'Non-elderly (<65 years) adults with asthma n= 28,912'
      [6]: 'Excluding 294 individuals with invalid \\n daily fruit or vegetable intake'
      [6]: 'Excluding 13,373 individuals with missing data \\n on variables used in the analysis'
      [8]: 'Final analytic sample \\n n= 15,245'
      ")
```





### Descriptive Statistics


#### Figure A. Histogram of Days of Poor Health among Non-elderly Adults with Asthma (BRFSS 2021)
```{r hist-poorHlth, warning=FALSE, message=FALSE}
brfss %>%
  ggplot(aes(x = days_poor_health)) +
  geom_histogram()
```


#### Figure B. Histogram of Days of Poor Health by Fruit Intake among Non-elderly Adults with Asthma (BRFSS 2021)
```{r hist-byFruit, warning=FALSE, message=FALSE}
brfss %>% 
  ggplot(aes(x = days_poor_health)) +
  geom_histogram() +
  facet_wrap(facets = vars(fruit_daily)) +
  labs(x = "Days of Poor Health", y = "Number of Participants") +
  theme_bw()

brfss %>% 
  ggplot(aes(x = days_poor_health)) +
  geom_histogram() +
  facet_wrap(facets = vars(veg_daily)) +
  labs(x = "Days of Poor Health", y = "Number of Participants") +
  theme_bw()
```


#### Figure C. Histogram of Days of Poor Health by Vegetable Intake among Non-elderly Adults with Asthma (BRFSS 2021)
```{r hist-byVeg, warning=FALSE, message=FALSE}
brfss %>% 
  ggplot(aes(x = days_poor_health)) +
  geom_histogram() +
  facet_wrap(facets = vars(veg_daily)) +
  labs(x = "Days of Poor Health", y = "Number of Participants") +
  theme_bw()
```


#### Table 1: Sample characteristics

```{r table1}
brfss %$%
  table1(~ age_cat + sex + race_eth + urban_rural + income + days_poor_health | fruit_daily, data=.)

brfss %$%
  table1(~ age_cat + sex + race_eth + urban_rural + income + days_poor_health | veg_daily, data=.)

brfss %$%
  table1(~ age_cat + sex + race_eth + urban_rural + income + fruit_daily + veg_daily + days_poor_health, data=.)
```


### Regression Models

#### Days of poor health by daily fruit

```{r uni-fruit}
# univariate poisson regression by daily fruit intake
mp_fruitOnly <- glm(days_poor_health ~ fruit_daily, family = poisson(), data = brfss)
summary(mp_fruitOnly)
#broom::tidy(mp_fruitOnly, exponentiate = TRUE, conf.int = TRUE)

# univariate negative binomial regression
mnb_fruitOnly <- glm.nb(days_poor_health ~ fruit_daily, data = brfss)
summary(mnb_fruitOnly)
broom::tidy(mnb_fruitOnly, exponentiate = TRUE, conf.int = TRUE)

lrtest(mp_fruitOnly, mnb_fruitOnly) # p-value is <0.05 so use negative binomial
```


```{r multi-fruit}
# multivariate poisson regression
mp_fruitFull <- glm(days_poor_health ~ fruit_daily + age_cat + sex + race_eth + urban_rural + income, family = "poisson", data = brfss)
summary(mp_fruitFull)
broom::tidy(mp_fruitFull, exponentiate = TRUE, conf.int = TRUE)

# multivariate negative binomial regression
mnb_fruitFull <- glm.nb(days_poor_health ~ fruit_daily + age_cat + sex + race_eth + urban_rural + income, data = brfss)
summary(mnb_fruitFull)
broom::tidy(mnb_fruitFull, exponentiate = TRUE, conf.int = TRUE)

lrtest(mp_fruitFull, mnb_fruitFull) # p-value is <0.05 so use negative binomial

car::vif(mnb_fruitFull)
```


```{r modelComp-fruit}
stargazer::stargazer(mp_fruitFull, mnb_fruitFull, title="Model Comparison", type="text",align=TRUE,single.row=TRUE, digits=6)
```


#### Fruit Interpretation

Adults with asthma who consumed fruit less than once per day had 1.02 (95% CI 0.97-1.08) times higher incidence of excess poor health days compared to adults with asthma who consumed fruit at least once daily, not a significant difference (adjusted for age, sex, race, urban/rural status and income).



#### Days of poor health by daily veg


```{r uni-veg}
# univariate poisson regression by daily veg intake
mp_vegOnly <- glm(days_poor_health ~ veg_daily, family = poisson(), data = brfss)
summary(mp_vegOnly)
#broom::tidy(mp_fruitOnly, exponentiate = TRUE, conf.int = TRUE)

# univariate negative binomial regression
mnb_vegOnly <- glm.nb(days_poor_health ~ veg_daily, data = brfss)
summary(mnb_vegOnly)
broom::tidy(mnb_vegOnly, exponentiate = TRUE, conf.int = TRUE)

lrtest(mp_vegOnly, mnb_vegOnly) # p-value is <0.05 so use negative binomial
```


```{r multi-veg}
# multivariate poisson regression
mp_vegFull <- glm(days_poor_health ~ veg_daily + age_cat + sex + race_eth + urban_rural + income, family = "poisson", data = brfss)
summary(mp_vegFull)
#broom::tidy(mp_vegFull, exponentiate = TRUE, conf.int = TRUE)

# multivariate negative binomial regression
mnb_vegFull <- glm.nb(days_poor_health ~ veg_daily + age_cat + sex + race_eth + urban_rural + income, data = brfss)
summary(mnb_vegFull)
broom::tidy(mnb_vegFull, exponentiate = TRUE, conf.int = TRUE)


lrtest(mp_vegFull, mnb_vegFull) # p-value is <0.05 so use negative binomial

car::vif(mnb_vegFull)
```


#### Veg Interpretation

Adults with asthma who consumed vegetables less than once per day had 1.08 (95% CI 1.003-1.16) times higher incidence of excess poor health days compared to adults with asthma who consumed vegetables at least once daily (adjusted for age, sex, race, urban/rural status and income).



## Sensitivity analysis (for personal reference, not included in proposal or abstract)

#### Check if poor health days coding issue (noted in line 201) affects results

```{r ALT-healthdays-recode}
brfss2021Clean_CHECK <- brfss2021 %>%
  # select variables of interest
  dplyr::select(ASTHNOW, POORHLTH, `_FRTLT1A`, `_VEGLT1A`, `_INCOMG1`, `_URBSTAT`, `_RACE`, `_SEX`, `_AGEG5YR`, PHYSHLTH, MENTHLTH, `_FRT16A`, `_VEG23A`) %>%
  # rename variables for ease of use
  rename(has_asthma = ASTHNOW) %>%
  rename(days_poor_health = POORHLTH) %>%
  rename(fruit_daily = `_FRTLT1A`) %>%
  rename(veg_daily = `_VEGLT1A`) %>%
  rename(income = `_INCOMG1`) %>%
  rename(urban_rural = `_URBSTAT`) %>%
  rename(race_eth = `_RACE`) %>%
  rename(sex = `_SEX`) %>%
  rename(age_cat = `_AGEG5YR`) %>%
  rename(days_physical_bad = PHYSHLTH) %>%
  rename(days_mental_bad = MENTHLTH) %>%
  rename(fruit_high = `_FRT16A`) %>%
  rename(veg_high = `_VEG23A`) %>%
  # recode numeric variable values
  mutate(days_poor_health = replace_na(days_poor_health, 0)) %>%
  mutate(days_poor_health = as.numeric(recode(as.character(days_poor_health), 
                                        '88' = '0', # none, 0 days poor health
                                        '77' = NA_character_, # don't know
                                        '99' = NA_character_))) %>% # refused
  mutate(days_physical_bad = as.numeric(recode(as.character(days_physical_bad), 
                                        '88' = '0', # none, 0 days poor health
                                        '77' = NA_character_, # don't know
                                        '99' = NA_character_))) %>% # refused
  mutate(days_mental_bad = as.numeric(recode(as.character(days_mental_bad), 
                                        '88' = '0', # none, 0 days poor health
                                        '77' = NA_character_, # don't know
                                        '99' = NA_character_))) %>% # refused
  # recode factor variable levels
  # for most variables, 9 = don't know, refused, missing
  ## for has_asthma, 7 = don't know, 9 = refused
  ## for age_cat, 14 = don't know, refused, missing
  mutate(has_asthma = recode_factor(has_asthma,
                                    '1' = "Has asthma",
                                    '2' = "No current asthma",
                                    '7' = NA_character_,
                                    '9' = NA_character_)) %>%
  mutate(fruit_daily = recode_factor(fruit_daily,
                                     '1' = "Consumed fruit at least once daily",
                                     '2' = "Consumed fruit < once daily",
                                     '9' = NA_character_)) %>%
  mutate(veg_daily = recode_factor(veg_daily,
                                     '1' = "Consumed vegetables at least once daily",
                                     '2' = "Consumed vegetables < once daily",
                                     '9' = NA_character_)) %>%
  mutate(income = recode_factor(income,
                                '1' = "Less than $15,000",
                                '2' = "$15,000 to < $25,000",
                                '3' = "$25,000 to < $35,000",
                                '4' = "$35,000 to < $50,000",
                                '5' = "$50,000 to < $100,000",
                                '6' = "$100,000 to < $200,000",
                                '7' = "$200,000 or more",
                                '9' = NA_character_)) %>%
  mutate(urban_rural = recode_factor(urban_rural,
                                     '1' = "Urban",
                                     '2' = "Rural")) %>%
  mutate(race_eth = recode_factor(race_eth,
                                '1' = "White non-Hispanic",
                                '2' = "Black non-Hispanic",
                                '3' = "American Indian / Alaskan Native non-Hispanic",
                                '4' = "Asian non-Hispanic",
                                '5' = "Native Hawaiian / Pacific Islander non-Hispanic",
                                '6' = "Other race non-Hispanic",
                                '7' = "Multiracial non-Hispanic",
                                '8' = "Hispanic",
                                '9' = NA_character_)) %>%
  mutate(age_cat = recode_factor(age_cat,
                                '1' = "18-24 years",
                                '2' = "25-29 years",
                                '3' = "30-34 years",
                                '4' = "35-39 years",
                                '5' = "40-44 years",
                                '6' = "45-49 years",
                                '7' = "50-54 years",
                                '8' = "55-59 years",
                                '9' = "60-64 years",
                                '10' = "65+ years",
                                '11' = "65+ years",
                                '12' = "65+ years",
                                '13' = "65+ years",
                                '14' = NA_character_)) %>%
  mutate(sex = recode_factor(sex,
                                     '1' = "Male",
                                     '2' = "Female"))

summary(brfss2021Clean_CHECK$days_poor_health)

brfssCleanFull_CHECK <- brfss2021Clean_CHECK %>%
  # include only those with current asthma
  filter(has_asthma == "Has asthma") %>%

  filter(age_cat != "65+ years") %>%

  filter(fruit_high==1, veg_high==1) %>%

  dplyr::select(-has_asthma, -fruit_high, -veg_high, -days_physical_bad, -days_mental_bad) %>%
  droplevels() %>%
  drop_na()
```


```{r ALT-healthdays-multiFruit}
mnb_fruitFull_CHECK <- glm.nb(days_poor_health ~ fruit_daily + age_cat + sex + race_eth + urban_rural + income, data = brfssCleanFull_CHECK)
summary(mnb_fruitFull_CHECK)
broom::tidy(mnb_fruitFull_CHECK, exponentiate = TRUE, conf.int = TRUE)
```


```{r ALT-healthdays-multiVeg}
# multivariate negative binomial regression
mnb_vegFull_CHECK <- glm.nb(days_poor_health ~ veg_daily + age_cat + sex + race_eth + urban_rural + income, data = brfssCleanFull_CHECK)
summary(mnb_vegFull_CHECK)
broom::tidy(mnb_vegFull_CHECK, exponentiate = TRUE, conf.int = TRUE)
```


